
---
title: "Course Project 1"
author: "Guillaume Polet"
date: "08/08/2018"
output:
  html_document: default
  pdf_document: default
---
## Instructions and Context
It is now possible to collect a large amount of data about personal movement using activity monitoring devices such as a Fitbit, Nike Fuelband, or Jawbone Up. These type of devices are part of the “quantified self” movement – a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. But these data remain under-utilized both because the raw data are hard to obtain and there is a lack of statistical methods and software for processing and interpreting the data.

This assignment makes use of data from a personal activity monitoring device. This device collects data at 5 minute intervals through out the day. The data consists of two months of data from an anonymous individual collected during the months of October and November, 2012 and include the number of steps taken in 5 minute intervals each day.

The data for this assignment can be downloaded from the course web site:


The variables included in this dataset are:

- steps: Number of steps taking in a 5-minute interval (missing values are coded as 
NA
NA)
- date: The date on which the measurement was taken in YYYY-MM-DD format
- interval: Identifier for the 5-minute interval in which measurement was taken

## Loading the Data and performong some preprocessing
```{r loaddata,warning=FALSE,echo=TRUE}
unzip(zipfile="activity.zip")  
data <- read.csv("activity.csv")
data$date <- as.POSIXct(data$date, "%Y-%m-%d")
weekday <- weekdays(data$date)
data <- cbind(data,weekday)

str(data)
summary(data)
```

## What is mean total number of steps taken per day?
```{r mean,warning=FALSE,echo=TRUE,message=FALSE}
if(require(ggplot2)==FALSE)(install.packages("ggplot2")); library(ggplot2)
if(require(dplyr)==FALSE)(install.packages("dplyr")); library(dplyr)
data2<-data%>%
        group_by(date)%>%
        summarise(activity_step=sum(steps,na.rm=TRUE))
ggplot(data=data2, aes(data2$activity_step)) + 
  geom_histogram( 
                 col="red", 
                 fill="green"
                 ) + 
  labs(title="Total number of steps taken per day")+
  xlab("Total steps taken per day")
  
mean(data2$activity_step,na.rm=TRUE)

median(data2$activity_step, na.rm=TRUE)

```

## What is the average daily activity pattern?
```{r averagedaily,warning=FALSE,echo=TRUE}
data3<-data%>%
  group_by(interval)%>%
  summarise(steps=mean(steps,na.rm=TRUE))

ggplot(data=data3, aes(x=interval, y=steps)) +
  geom_line() +
  xlab("5-minute interval") +
  ylab("average number of steps taken")
```

On average across all the days in the dataset, the 5-minute interval contains
the maximum number of steps?
```{r maxavg,warning=FALSE,echo=TRUE}
data3[which.max(data3$steps), ]$interval
```

## Imputing missing values

There are many days/intervals where there are missing values (coded as `NA`). The presence of missing days may introduce bias into some calculations or summaries of the data.

```{r missing,warning=FALSE,echo=TRUE}
## missing
sum(is.na(data$steps))



```

All of the missing values should be imputed with mean value for that 5-minute
interval.

```{r imputation,warning=FALSE,echo=TRUE}
# Creating a function to imput missing values with mean for that 5 minute interval
valueImputation <- function(steps, interval) {
  filled <- NA
  if (!is.na(steps))
    filled <- c(steps)
  else
    filled <- (data3[data3$interval==interval, "steps"])
  return(filled)
}
imputedData<- data
imputedData$steps <- mapply(valueImputation, imputedData$steps, imputedData$interval)

```
Now, using the filled data set, let's make a histogram of the total number of steps taken each day and calculate the mean and median total number of steps.

```{r imputeddata,warning=FALSE,echo=TRUE}
data5<-imputedData%>%
  group_by(date)%>%
  summarise(activity_step=sum(as.numeric(steps),na.rm=TRUE))

ggplot(data=data5, aes(data5$activity_step)) + 
  geom_histogram( 
    col="red", 
    fill="green") + 
  labs(title="Total number of steps taken per day")+
  xlab("Total steps taken per day")
```
Here are the mean and the median
```{r,warning=FALSE,echo=TRUE}
mean(data5$activity_step,na.rm=TRUE)

median(data5$activity_step, na.rm=TRUE)
```
Mean and median values are higher after imputing missing data. The reason is
that in the original data, there are some days with `steps` values `NA` for 
any `interval`. The total number of steps taken in such days are set to 0s by
default. However, after replacing missing `steps` values with the mean `steps`
of associated `interval` value, these 0 values are removed from the histogram
of total number of steps taken each day.

## Are there differences in activity patterns between weekdays and weekends?
Create a new factor variable in the dataset with two levels -- "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.

```{r,warning=FALSE,echo=TRUE,message=FALSE}
dayOrWeeekend <- function(date) {
  day <- weekdays(date)
  if (day %in% c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday"))
    return("weekday")
  else if (day %in% c("Saturday", "Sunday"))
    return("weekend")
  else
    stop("invalid date")
}
imputedData$date <- as.Date(imputedData$date)
imputedData$day <- sapply(imputedData$date, FUN=dayOrWeeekend)
```

Make a panel plot containing a time series plot (i.e. type = “l”) of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). See the README file in the GitHub repository to see an example of what this plot should look like using simulated data.
```{r avgweekdayvsweekend,warning=FALSE,echo=TRUE,message=FALSE}
data6<-imputedData%>%
  group_by(interval,day)%>%
  summarise(steps=mean(as.numeric(steps),na.rm=TRUE))

ggplot(data6, aes(interval, steps,color=day)) +
geom_line() +
facet_grid(day ~ .)+ 
xlab("5-minute interval") +
  ylab("Number of steps")+
  theme(legend.position="none")
```

